Statically Typed String Sanitation Inside a Python

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Updated Sec 2.1.4

<$e$> ::= 
      $x$ | $\lambda x . e$ | $e(e) $\hfill source terms \alt 
      $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
      $\sisubst{r}{e}{e}$ | $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$

@@

\inferrule[S-T-Replace]
{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}

@@

\inferrule[S-T-Check]
{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
\end{mathpar}

@@

\inferrule[S-E-Replace]
{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}

@@

\inferrule[S-E-Check-NotOk]
{
  e \sreduces \strin{s} \\ s \not \in \lang{r} \\
  e_2 \sreduces v
}
{
  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
}
\end{mathpar}
\caption{Big step semantics for $\lambdas$}.
\label{fig:dlambdas}
\end{figure}

@@

\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then

  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{prop}
\begin{proof}
The proposition follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
\end{proof}

@@

In this section, we establish type soundness for $\lambdas$.
The theorem relies upon the lemmas and propositions established above.
We also rely on additional lemmas which establish the preservation of well-formedness
of regular expressions.

\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
The following are properties of regular expressions which are necessary for our proofs:
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
Regular languages are closed under reversal.
\end{lem}

@@

\begin{lem}
  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}
\begin{proof}
  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
\end{proof}

Safety for the string fragment of $\lambdas$ requires validating that the type system's definition is justified
by our theorems about regular languages.
We avoid the most significant issues inherent to big-step semantics by avoiding non-termination.
The simply typed lambda calculus terminates, and our conservative, compositional extension clearly
terminates modulo termination of ${\sf subst}$ and ${\sf lreplace}$ (but even if these were non-terminating, this still does not pose a problem for the big-step semantics itself).
However, a more careful treatment or treatment for a language with non-termination might proceed by either coinduction \cite{coinductionoo, coinduction} or step-indexing \cite{stepindexing}.

\begin{thm}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
  If $\sctx \vdash e : \stringin{r}$ 
  then $e \sreduces \rstr{s}$ 
\end{thm}

\begin{thm}[Type Safety.] \label{thm:scorrect}
  If $\sctx \vdash e : \sigma$ 
  then $e \sreduces v$ and $\sctx \vdash v : \sigma$.
\end{thm}
\begin{proof}
By induction on the typing relation.
The S-T-Concat case requires Lemma \ref{thm:regexprops} 
and the S-T-Replace case appeals to Lemma \ref{thm:substcorrespondence}.
\end{proof}

@@

The chief benefit of $\lambdas$ is its security theorem, which states that any value of
a regular expression type is recognized by the regular expression corresponding to its type.
This is beneficial because safe subsets of strings may be characterized by regular expressions.
Our main technical result, stated later in this section, establishes that the security property
is preserved under translation into $\lambdap$.

\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
\end{thm}
\begin{proof}
  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
\end{proof}

@@

\end{grammar}
\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}

The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
with a $\str$ type and a regular expression type, as well as 
some operations -- such as concatenation and replacement -- found in many the standard libraries of many programming languages.
The operations of $\lambdap $ correspond precisely to the operations of $\lambdas$ in a way made precise by the translation rules described in the next section.
Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
such as \textbf{P}ython or \textbf{P}HP.

>>>>

<$e$> ::= 
      $x$ | $\lambda x . e$ | $e(e) $\hfill source terms \alt 
      $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$ | $\sisubst{r}{e}{e}$

@@

@@

\inferrule[S-T-Check]
{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}

\inferrule[S-T-Replace]
{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
\end{mathpar}

@@

@@

\inferrule[S-E-Check-NotOk]
{
  e \sreduces \strin{s} \\ s \not \in \lang{r} \\
  e_2 \sreduces v
}
{
  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
}

\inferrule[S-E-Replace]
{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
\end{mathpar}
\caption{Big step semantics for $\lambdas$}.
\label{fig:dlambdas}
\end{figure}

@@

\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{prop}
\begin{proof}
The proposition follows from the definitions of $\textsf{subst}$ and $\textsf{lreplace}$; note that language substitutions over-approximate string substitutions.\todo{elaborate on this?}
\end{proof}

@@

In this section, we establish type soundness for $\lambdas$.
The theorem relies upon the definitions and propositions given above and some basic properties of regular languages.

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}
\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item Regular languages are closed under reversal.
\end{enumerate}
\end{lem}

@@

We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
\begin{lem}
  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}
\begin{proof}
  The only non-trivial case is S-T-Replace, which follows from Proposition \ref{thm:total}. \todo{what about lhead/ltail?}
\end{proof}

Safety for the string fragment of $\lambdas$ requires validating that the type system's definition is justified
by our theorems about regular languages.

We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
The simply typed lambda calculus terminates, and our conservative extension clearly
terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety holds.
However, a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.

\begin{thm}[Type Safety.] \label{thm:scorrect}
  If $\emptyset \vdash e : \sigma$ 
  then $e \sreduces v$ and $\emptyset \vdash v : \sigma$.
\end{thm}
\begin{proof}
By induction on the typing relation.
The \textsc{S-T-Concat} case requires Lemma \ref{thm:regexprops}.1, the \textsc{S-T-Check} case requires Lemma \ref{thm:regexprops}.2 
and the S-T-Replace case appeals to Proposition \ref{thm:substcorrespondence}.
\end{proof}

We can also define a canonical forms lemma for regular strings.

\begin{lemma}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lemma}
\begin{proof}
The proof is again by induction on the typing relation, following straightforwardly.
\end{proof}

@@

The chief benefit of $\lambdas$ is its security theorem, which states that any value of
a regular string type is recognized by the regular language corresponding to the regular expression the type is indexed by.
This is beneficial because this ensure that membership in a regular language known to be secure becomes manifest in the type of the string, rather than being a property that must be established extralinguistically.

\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
  If  $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
\end{thm}
\begin{proof}
  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of \textsc{S-T-Stringin-I}, the typing rule for terms of the form $\rstr{s}$.
\end{proof}

@@

\end{grammar}
\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}
Our next major technical result, stated in Sec. \ref{sec:tr}, establishes that the security property
is preserved under translation into $\lambdap$, which we now define.

The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
with a $\str$ type and a $\regex$ type, as well as 
some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section.
%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
such as \textbf{P}ython or \textbf{P}HP. We will discuss an implementation within the former in the next section.
